---
title: "Ancient Mediterranean Shipwrecks"
author: "ADS"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE
  )

library(dplyr)
library(ggplot2)
library(ggspatial)
library(readr)
library(rnaturalearth)
library(rnaturalearthdata)
library(rosm)
library(sf)
library(viridis)

shipwrecks <- read_csv("processed_data/shipwrecks.csv")

```

The aim for this challenge is to use a dataset available on the internet to attempt to recreate
an often-cited graph in Peregrine Horden and Nicholas Purcell's book, *The Corrupting Sea*.

The graph is part of a larger argument in which Horden and Purcell hold out the possibility that
the volume of sea-borne trade in the Mediterranean was larger in the 2nd century CE than at any
other time before the 19th century.

# The Challenge

Download a copy of the dataset cited below, import the data, and carry out the following tasks.

> Strauss, Julia. 2013. ‘Shipwrecks Database’. *The Oxford Roman Economy Project*. \
> oxrep.classics.ox.ac.uk/databases/shipwrecks_database/.

## Dates

The date_range and mid_point_of_date_range variables contain only zeros.

1. Calculate the mid-point and fill the `mid_point_of_date_range` column with data.
2. Calculate the date range (which amounts to a measure of the uncertainty of the dating) and 
   fill the `date_range` column with data. Remember that the dates include negative values (BCE).

```{r shipwrecks-dates}

date_range <- function(earliest, latest) {
  abs(latest) - abs(earliest)
}


shipwrecks_dates <- shipwrecks %>% 
  mutate(
    date_range = date_range(earliest_date, latest_date)
  ) %>% 
  mutate(
    mid_point_of_date_range = round((earliest_date + latest_date) / 2, 0)
  )

```

Using the updated data, answer these questions:

1. How many shipwrecks are there in the data?
1. What is the earliest shipwreck?
2. What is the latest shipwreck?

```{r}
shipwrecks_dates %>% 
  summarise(
    count = n(),
    earliest = min(mid_point_of_date_range, na.rm = TRUE),
    latest = max(mid_point_of_date_range, na.rm = TRUE)
  )
```


## Locations

1. What countries have the shipwrecks been found in?
2. How many wrecks have been discovered in each country?
3. How many shipwrecks in the database have a location specified by a valid lat-long pair?


### What countries?

```{r}
wreck_countries <- unique(shipwrecks_dates$country)

wreck_countries
```


## Shipwrecks with valid lat-long pair?

```{r valid-lat-long-count}

shipwrecks_dates %>% 
  filter(
    !is.na(latitude) | !is.na(longitude)
  ) %>% 
  filter(
    !(latitude == 0 & longitude == 0)
  ) %>% 
  summarise(
    count = n(),
    earliest = min(mid_point_of_date_range, na.rm = TRUE),
    latest = max(mid_point_of_date_range, na.rm = TRUE),
    min_lat = min(latitude, na.rm = TRUE),
    max_lat = max(latitude, na.rm = TRUE),
    min_long = min(longitude, na.rm = TRUE),
    max_long = max(longitude, na.rm = TRUE)
  )

```


## Bonus challenge

1. Plot the shipwrecks that have valid long-lat coordinates onto a map (use `ggplot` and 
   `rnaturalearth` for the maps).
      - Read the `ggplot2` in-progress book (3rd edition) about mapping: \
         [https://ggplot2-book.org/maps.html](https://ggplot2-book.org/maps.html).
      - Use these snippets to add a map layer from `rnaturalearth`:
        - `world <- ne_countries(scale = "medium", returnclass = "sf")`
        - `ggplot(data = world) + geom_sf()`
2. Create a faceted set of maps on some variable of your choice, e.g. date ranges.

```{r plot_shipwrecks, fig.height=8}
# get data for labelling individual countries
countries <- read_tsv("processed_data/countries.tsv") %>% 
  # only relevant countries for the shipwreck data
  filter(
    name %in% unique(shipwrecks_dates$country)
  ) %>% 
  # avoid name clash
  select(-name)
# add labelling data onto the Natural Earth countries data
world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
  left_join(countries, by = c("iso_a2" = "country"))

shipwrecks_dates_locations <- shipwrecks_dates %>%
  filter(
    !is.na(latitude) | !is.na(longitude)
  ) %>% 
  filter(
    !(latitude == 0 & longitude == 0)
  )

# make the plot
ggplot(data = world) +
  # set up the world map with NE countries data
  geom_sf() +
  # layer on points for the shipwrecks
  geom_point(
    data = shipwrecks_dates_locations, aes(
      x = longitude, 
      y = latitude, 
      colour = mid_point_of_date_range,
    ), 
    alpha = 0.75, size = 3
  ) +
  # add labels for relevant countries
  geom_label(
    data = world, aes(x = longitude, y = latitude, label = name), nudge_y = 0.3, size = 3
  ) +
  # crop the map to the Mediterranean
  coord_sf(xlim = c(-1, 35), ylim = c(30, 47), expand = TRUE) +
  # select an appropriate colour scale
  scale_color_binned(name = "Approx. date", type = "viridis", option = "plasma") +
  # adjust the overall visual appearance
  theme_bw() +
  theme(
    panel.grid.major = element_line(color = gray(0.5), linetype = "dashed", size = 0.5),
    panel.background = element_rect(fill = "lightsteelblue1")
  ) +
  # add labels for the overall map
  labs(
    title = "Ancient Mediterranean shipwrecks",
    subtitle = "358 wrecks with known location data, 1800 BCE to 1500 CE",
    caption = "Data from Strauss, Julia. 2013. ‘Shipwrecks Database’. The Oxford Roman Economy Project. oxrep.classics.ox.ac.uk/databases/shipwrecks_database/."
  ) +
  # add direction and scale annotations from the package `ggspatial`
  annotation_scale(location = "bl", width_hint = 0.4) +
  annotation_north_arrow(
    location = "bl", 
    which_north = "true",
    pad_x = unit(1, "cm"),
    pad_y = unit(1, "cm")
  )

ggsave("output/shipwrecks_plot.png", device = png)
```

